---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# subsampling

<!-- badges: start -->
[![R-CMD-check](https://github.com/dqksnow/Subsampling/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/dqksnow/Subsampling/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The R package `subsampling` provides optimal subsampling methods for common statistical models such as glm, softmax(multinomial) regression, rare event logistic regression and quantile regression model.

## Installation

You can install the development version of subsampling from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("dqksnow/Subsampling")
```

## Getting Started

The [Online document]() provides a guidance for quick start.

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(subsampling)
set.seed(1)
N <- 5e3
beta0 <- rep(-0.5, 7)
d <- length(beta0) - 1
corr <- 0.5
sigmax  <- matrix(corr, d, d) + diag(1-corr, d)
X <- MASS::mvrnorm(N, rep(0, d), sigmax)
colnames(X) <- paste("V", 1:ncol(X), sep = "")
P <- 1 - 1 / (1 + exp(beta0[1] + X %*% beta0[-1]))
Y <- rbinom(N, 1, P)
data <- as.data.frame(cbind(Y, X))
formula <- Y ~ .
n.plt <- 100
n.ssp <- 500
ssp.results <- ssp.glm(formula = formula,
                       data = data,
                       n.plt = n.plt,
                       n.ssp = n.ssp,
                       family = "quasibinomial",
                       criterion = "optL",
                       sampling.method = "poisson",
                       likelihood = "weighted"
                       )
summary(ssp.results)
```
