---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# subsampling

<!-- badges: start -->
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/subsampling)](https://CRAN.R-project.org/package=subsampling)
[![Total_Downloads](https://cranlogs.r-pkg.org/badges/grand-total/subsampling)](https://CRAN.R-project.org/package=subsampling)
[![Monthly_Downloads](https://cranlogs.r-pkg.org/badges/subsampling)](https://CRAN.R-project.org/package=subsampling)
[![R-CMD-check](https://github.com/dqksnow/subsampling/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/dqksnow/subsampling/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

A major challenge in big data statistical analysis is the demand for computing
resources. For example, when fitting a logistic regression model to binary
response variable with $N \times d$ dimensional covariates, the computational
complexity of estimating the coefficients using the IRLS
algorithm is $O(\zeta N d^2)$, where $\zeta$ is the number of iteriation. When $N$ is large, the cost can be
prohibitive, especially if high performance computing resources are
unavailable. Subsampling has become a widely used
technique to balance the trade-off between computational efficiency and statistical
efficiency.

The R package `subsampling` provides optimal subsampling methods for various statistical models such as generalized linear models (GLM), softmax (multinomial) regression, rare event logistic regression and quantile regression model. Specialized subsampling techniques are provided to address specific challenges across different models and datasets. With specified model assumptions and subsampling techniques, it draws subsample from the full data, fits model on the subsample and perform statistical inferences.

## Installation

You can install the package by

``` r
# Install from CRAN
install.packages("subsampling")

# Or install the development version from GitHub
# install.packages("devtools")
devtools::install_github("dqksnow/subsampling")
```

## Getting Started

The [Online document](https://dqksnow.github.io/subsampling/) provides a guidance for quick start.

- [Generalized linear model](https://dqksnow.github.io/subsampling/articles/ssp-logit.html).
- [Rare event logistic regression](https://dqksnow.github.io/subsampling/articles/ssp-relogit.html).
- [Softmax (multinomial) regression](https://dqksnow.github.io/subsampling/articles/ssp-softmax.html).
- [Quantile regression](https://dqksnow.github.io/subsampling/articles/ssp-quantreg.html).
  
## Example

This is an example of subsampling method on logistic regression:

```{r example}
library(subsampling)
set.seed(1)
N <- 1e4
beta0 <- rep(-0.5, 7)
d <- length(beta0) - 1
corr <- 0.5
sigmax  <- matrix(corr, d, d) + diag(1-corr, d)
X <- MASS::mvrnorm(N, rep(0, d), sigmax)
colnames(X) <- paste("V", 1:ncol(X), sep = "")
P <- 1 - 1 / (1 + exp(beta0[1] + X %*% beta0[-1]))
Y <- rbinom(N, 1, P)
data <- as.data.frame(cbind(Y, X))
formula <- Y ~ .
n.plt <- 200
n.ssp <- 600
ssp.results <- ssp.glm(formula = formula,
                       data = data,
                       n.plt = n.plt,
                       n.ssp = n.ssp,
                       family = "quasibinomial",
                       criterion = "optL",
                       sampling.method = "poisson",
                       likelihood = "weighted"
                       )
summary(ssp.results)
```

## Acknowledgments

The development of this package was supported by the National Eye Institute of the National Institutes of Health under Award Number R21EY035710.