---
title: "Introduction to `ssp.glm`: Subsampling for Generalized Linear Models"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Introduction to `ssp.glm`: Subsampling for Generalized Linear Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette introduces the usage of the `ssp.glm` using logistic regression as
an example of generalized linear models (GLM). The statistical theory and
algorithms in this implementation can be found in the relevant reference papers.

The log-likelihood function for a GLM is

$$
\max_{\beta} L(\beta) = \frac{1}{N} \sum_{i=1}^N \left\{y_i u(\beta^{\top} x_i)
- \psi \left[ u(\beta^{\top} x_i) \right] \right\},
$$
where $u$ and $\psi$ are known functions depend on the distribution from the
exponential family. For the binomial distribution, the log-likelihood function
becomes

$$
\max_{\beta} L(\beta) = \frac{1}{N} \sum_{i=1}^N \left[y_i \beta^{\top} x_i -
\log\left(1 + e^{\beta^\top x_i}\right) \right].
$$

The idea of subsampling methods is as follows: instead of fitting the model on
the size $N$ full dataset, a subsampling probability is assigned to each
observation and a smaller, informative subsample is drawn. The model is then
fitted on the subsample to obtain an estimator with reduced computational cost.

## Installation

You can install the development version of subsampling from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("dqksnow/Subsampling")
```

```{r setup}
library(subsampling)
```

## Terminology

- Full dataset: The whole dataset used as input.

- Full data estimator: The estimator obtained by fitting the model on the full
  dataset.

- Subsample: A subset of observations drawn from the full dataset.

- Subsample estimator: The estimator obtained by fitting the model on the
  subsample.

- Subsampling probability ($\pi$): The probability assigned to each observation
  for inclusion in the subsample.

## Example: Logistic Regression with Simulated Data

We introduce the usage of `ssp.glm` with simulated data. $X$ contains $d=6$
covariates drawn from multinormal distribution, and $Y$ is the binary response
variable. The full dataset size is $N = 1 \times 10^4$.

```{r}
set.seed(1)
N <- 1e4
beta0 <- rep(-0.5, 7)
d <- length(beta0) - 1
corr <- 0.5
sigmax  <- matrix(corr, d, d) + diag(1-corr, d)
X <- MASS::mvrnorm(N, rep(0, d), sigmax)
colnames(X) <- paste("V", 1:ncol(X), sep = "")
P <- 1 - 1 / (1 + exp(beta0[1] + X %*% beta0[-1]))
Y <- rbinom(N, 1, P)
data <- as.data.frame(cbind(Y, X))
formula <- Y ~ .
head(data)
```

The function usage is

```{r, eval = FALSE}
ssp.glm(
  formula,
  data,
  subset = NULL,
  n.plt,
  n.ssp,
  family = "quasibinomial",
  criterion = "optL",
  sampling.method = "poisson",
  likelihood = "weighted",
  control = list(...),
  contrasts = NULL,
  ...
  )
```

## Key Arguments

The core functionality of `ssp.glm` revolves around three key questions:

- How are subsampling probabilities computed? (Controlled by the `criterion`
  argument)

- How is the subsample drawn? (Controlled by the `sampling.method` argument)

- How is the likelihood adjusted to correct for bias? (Controlled by the
  `likelihood` argument)

### `criterion`

The choices of `criterion` include `optA`, `optL`(default), `LCC` and
`uniform`. The optimal subsampling criterion `optA` and `optL` are derived by
minimizing the asymptotic covariance of subsample estimator, proposed by
@wang2018optimal. `LCC` and `uniform` are baseline methods.

### `sampling.method`

The options for the `sampling.method` argument include `withReplacement` and
`poisson` (default). `withReplacement` stands for drawing `n.ssp` subsamples
from full dataset with replacement, using the specified subsampling
probability. `poisson` stands for drawing subsamples one by one by comparing the
subsampling probability with a realization of uniform random variable
$U(0,1)$. The expected number of drawn samples are `n.ssp`. More details see
@wang2019more.

### `likelihood`

The available choices for `likelihood` include `weighted` (default) and
`logOddsCorrection`. Both of these likelihood functions can derive an unbiased
estimator. Theoretical results indicate that `logOddsCorrection` is more
efficient than `weighted` in the context of logistic regression. See
@wang2022maximum.

## Results

After drawing subsample, `ssp.glm` utilizes `survey::svyglm` to fit the model on
the subsample, which eventually uses `glm`. Arguments accepted by `svyglm` can
be passed through `...` in `ssp.glm`.

Below are two examples demonstrating the use of `ssp.glm` with different
configurations.

```{r}
n.plt <- 200
n.ssp <- 600
ssp.results <- ssp.glm(formula = formula,
                       data = data,
                       n.plt = n.plt,
                       n.ssp = n.ssp,
                       family = "quasibinomial",
                       criterion = "optL",
                       sampling.method = "withReplacement",
                       likelihood = "weighted"
                       )
summary(ssp.results)
```

```{r}
ssp.results <- ssp.glm(formula = formula,
                       data = data,
                       n.plt = n.plt,
                       n.ssp = n.ssp,
                       family = "quasibinomial",
                       criterion = "optA",
                       sampling.method = "poisson",
                       likelihood = "logOddsCorrection"
                       )
summary(ssp.results)
```

As recommended by `survey::svyglm`, when working with binomial models, it is
advisable to use use `family=quasibinomial()` to avoid a warning issued by
`glm`. Refer to [svyglm() help documentation Details
](https://www.rdocumentation.org/packages/survey/versions/4.4-2/topics/svyglm). The
'quasi' version of the family objects provide the same point estimates.

### Returned object 

The object returned by `ssp.glm` contains estimation results and indices of the
drawn subsample in the full dataset.

```{r}
names(ssp.results)
```

Some key returned variables:

- `index.plt` and `index` are the row indices of drawn pilot subsamples and
optimal subsamples in the full data. They are ready to be used for further
analysis or downstream tasks.

- `coef.ssp` is the subsample estimator for $\beta$ and `coef` is the linear 
combination of `coef.plt` (pilot estimator) and `coef.ssp`. 

- `cov.ssp` and `cov` are estimated covariance matrices of `coef.ssp` and
`coef`. 

The coefficients and standard errors printed by `summary()` are `coef` and the
square root of `diag(cov)`. See the help documentation of `ssp.glm` for details.

## Other Families

We also provide examples for poisson regression and gamma regression in the help
documentation of `ssp.glm`. Note that `likelihood = logOddsCorrection` is
currently implemented only for logistic regression (family = binomial or
quasibonomial).

### References